Inferring Continuous and Discrete Population Genetic Structure Across Space

### 2017-07-27 writeup/conStruct.tex
G's edits

\begin{equation}
\widehat{\Omega} = \frac{1}{L}  FF^T
\label{sample_covariance}
\end{equation}

\paragraph{Continuous and discrete differentiation}
We wish to describe the observed patterns of genetic variation as a combination of 
discrete population clusters,
within which genetic variation is continuously distributed, 
and between which sampled individuals or populations can be admixed.
Each of these discrete clusters is modeled as a spatial process
in which migration between neighboring demes acts 
to homogenize allele frequency changes that arise locally due to drift;
this interplay between drift and migration 
results in a continuous pattern of isolation by distance within each cluster.
Multiple population cluster processes can co-occur in space, 
leading to discrete jumps in allele frequencies across many loci over small geographic distances,
and the genotyped samples can be admixed between different clusters.

The continuous decay of allele frequency covariance with geographic distance 
within a cluster is described using a simple and flexible powered exponential function;
the within-cluster covariance between samples $i$ and $j$ is given by:

@@

where $G_{i,j}$ is the covariance function between samples $i$ and $j$ within a cluster,
and the superscript $(k)$ denotes that this function is specific to the $k$th cluster.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
the $\alpha$ parameters control the shape of the decay of covariance with distance,
and $\mu$ is a parameter that describes the amount of shared drift within a cluster.
The shared drift parameter $\mu^{(k)}$ can be interpreted as the branch length connecting 
the $k$th population to the population ancestral to all modeled clusters.

Genotyped samples can be admixed between these different clusters. 
The admixture proportion of the $i$th sample in the $k$th cluster, $w^{(k)}_i$,
gives the probability that an allele in sample $i$ was derived from cluster $k$;
each sample's admixture proportions across all clusters must sum to one.

We can then describe the covariance between samples $i$ and $j$ across all clusters, $\Omega_{i,j}$,
by summing their within-cluster spatial covariances ($G_{i,j}^{(k)}$ in cluster $k$) across all $K$ clusters
and weighting by the product of those samples' admixture proportions in each cluster.

\begin{equation}
\Omega_{i,j} = \gamma + \sum\limits_K w^{(k)}_i w^{(k)}_j G^{(k)}_{i,j} + \delta_{i,j}\eta_i
\label{cross_cluster_covariance}
\end{equation}

In addition to the admixture-weighted sum of the within-cluster spatial covariances,
this function contains two terms, $\gamma$ and $\delta_{i,j}\eta_i$.
The first, $\gamma$, describes the global covariance between all samples, 
and arises because all samples share an ancestral mean allele frequency at each locus.
In the second term, $\delta_{i,j}$ is an indicator variable that takes a value of 1 when $i$ equals $j$ and 0 otherwise. 
and $\eta_i$ gives the variance specific to sample $i$, 
This term on the diagonal of the parametric covariance matrix 
is meant to capture processes shaping variance within the sampled deme, 
such as inbreeding.

\paragraph{Likelihood and inference}
If we assume that the  allele frequencies $\widehat{F}$ are independent and multivariate normally distributed, 
their sample covariance $\widehat{\Omega}$ 
will be Wishart distributed with degrees of freedom equal to $L$, 
the number of loci genotyped.
Linkage disequilibrium (LD) between loci will decrease the effective number of degrees of freedom, 
which we discuss further below.
The likelihood of the sample standardized allele frequency covariance is therefore given by
\begin{equation}
P(\widehat{\Omega} \; | \; \Omega) = \mathcal{W} \left( L\widehat{\Omega} \; | \; \Omega,L\right)
\end{equation}

@@

>>>>

\begin{equation}
\widehat{\Omega} = \frac{1}{L}  FF^T
\label{sample_covariance}
\end{equation}
\gc{SO IS THIS ACTUALLY the calc given the new diagonal?}

\paragraph{Continuous and discrete differentiation}
We wish to describe the observed patterns of genetic variation as a combination of 
discrete population clusters,
within which genetic variation is continuously distributed, 
and between which sampled individuals or populations can be admixed.
Each of these discrete clusters is modeled as a spatial process where
\gc{allele frequencies covary between geographically close locations,
  but  %%changed as we dont model migration or drif
this covariance is allowed to decay with geographic distance. This 
pattern of spatial decay reflects} how migration between neighboring demes acts 
to homogenize allele frequency changes that arise locally due to
drift, \gc{but that migration is less effective at homogenizing
  between geographic distance demes};
%with this interplay between drift and migration 
resulting in a continuous pattern of isolation by distance within each cluster.
Multiple population cluster processes can co-occur in space, 
leading to discrete jumps in allele frequencies across many loci over small geographic distances,
and the genotyped samples can be admixed between different clusters.

\gc{To model the} continuous decay of allele frequency covariance with geographic distance 
within a cluster \gc{we use a flexible powered-exponential function;
  where the} within-cluster covariance between samples $i$ and $j$ is given by:

@@

where $G_{i,j}$ is the covariance function between samples $i$ and $j$ within a cluster,
and the superscript $(k)$ denotes that this function is specific to the $k$th cluster.
The quantity $D_{i,j}$ is the observed geographic distance between samples $i$ and $j$,
the \gc{$\alpha^{(k)}$ parameters control the shape of the decay of
  covariance with distance in the cluster},
and $\mu$ is a parameter that describes the \gc{background allele
  frequency   %%amount of shared drift
  covariance within the cluster. If two samples draw 100\% of their
  ancestry from cluster $k$ then, if they are geographically very
  close ($D_{i,j}=0$) they will have allele frequency covariance
  $\alpha^{(k)}_0 +  \mu^{(k)}$ with their covariance decaying to the
  background level $\mu^{(k)}$ within the cluster for large enough geographic distances.}
The shared drift parameter $\mu^{(k)}$ can be interpreted as the branch length connecting 
the $k$th population to the population ancestral to all modeled
clusters \gc{(see XXX)}.

\gc{Genotyped samples can draw their ancestry from multiple different clusters,
representing these samples being admixed between the clusters.} 
The admixture proportion of the $i$th sample in the $k$th cluster, $w^{(k)}_i$,
gives the probability that an allele in sample $i$ was derived from
cluster $k$ \gc{($\sum_k w^{(k)}_i =1$).}

We can then describe the covariance between samples $i$ and $j$ across all clusters, $\Omega_{i,j}$,
by summing their within-cluster spatial covariances ($G_{i,j}^{(k)}$ in cluster $k$) across all $K$ clusters
and weighting by the product of those samples' admixture proportions in each cluster
\begin{equation}
\Omega_{i,j} = \gamma + \sum\limits_K w^{(k)}_i w^{(k)}_j
G^{(k)}_{i,j} + \delta_{i,j}\eta_i .
\label{cross_cluster_covariance}
\end{equation}

In addition to the admixture-weighted sum of the within-cluster spatial covariances,
this function contains two terms, $\gamma$ and $\delta_{i,j}\eta_i$.
The first, $\gamma$, describes the global covariance between all samples, 
and arises because all samples \gc{have a base-line covariance as they
  share} an ancestral mean allele frequency at each locus.
In the second term, $\delta_{i,j}$ is an indicator variable that takes a value of 1 when $i$ equals $j$ and 0 otherwise. 
and $\eta_i$ gives the variance specific to sample $i$, 
This term on the diagonal of the parametric covariance matrix \gc{captures} processes shaping variance within the sampled deme, 
such as inbreeding \gc{and the sampling process}.

\paragraph{Likelihood and inference}
\gc{We} assume that the allele frequencies \gc{at each locus
  ($\widehat{F}_{\ell}$) are independent among loci and multivariate normally
  distributed across populations}, 
their sample covariance $\widehat{\Omega}$ 
will be Wishart distributed with degrees of freedom equal to $L$, 
the number of loci genotyped.
Linkage disequilibrium (LD) between loci will decrease the effective number of degrees of freedom, 
which we discuss further below.
The likelihood of the sample standardized allele frequency covariance is therefore given by
\begin{equation}
P(\widehat{\Omega} \; | \; \Omega) = \mathcal{W} \left( L\widehat{\Omega} \; | \; \Omega,L\right)
\end{equation}

@@

\gc{
\paragraph{Relationship of this model to non-spatial structure models}

}

\gc{ 
\paragraph{Choice of cluster number and cross-validation}

}
